|

### 2019-09-22 paper.Rmd
Spellcheck

<!-- Background -->
Data is commonly high-dimensional, and visualisation often relies on some form of dimension reduction. This can be done by taking linear projections, or nonlinear if one considers techniques like multidimensional scaling (MDS) [@mds] or t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne]. For the purposes here, the focus is on linear projections, in particular as provided by the grand tour [@As85; @BCAH05]. Interactive and dynamic displays can provide information beyond what can be achieved in a static display. The grand tour shows a smooth sequence of interpolated low dimensional projections, and allows the viewer to extrapolate from the low-dimensional shapes to the multidimensional distribution. It is particularly useful for detecting clusters, outliers and non-linear dependence.

<!-- The disadvantage of projection: opacity, occlusion, even with pseudo-transparency from scatterplots;  motivation -->
A major limitation of projections is their opacity. It is mitigated some, when scatterplots are used to render the projected data, through a pseudo-transparency of sparseness of points. This is improved further if points are also drawn using an alpha level that provides transparent dots on most display devices. Some features of multivariate distributions may be visible, but a lot is easily hidden, especially in the case of concave structures. Considering the example of a simple geometric shape such as a hypershere, it is difficult to distinguish between a full or a hollow sphere, that is, whether points are uniformly distributed within the sphere or on the surface of a sphere. Similarly we can think of small scale structures hidden in the center of a multivariate distribution, that might be considered to be "needles in a haystack", which can be difficult to detect in projections. Projections also obscure non-linear boundaries as might be constructed from a classification model, or nonlinear model fits in high-dimensions.

@@

For the explanation of the slice tour, the actual mechanics of the tour are not important, and only the notion of a projection plane in high dimensions is needed. The notation used for explanations starts with deoting each projection as $\mathbf{X}$ the $n\times p$ dimensional data matrix ($n$ observations in $p$ dimensions) and $\mathbf{A}$ an orthonormal $p \times d$ projection matrix. The $d$-dimensional projection of the data is thus given by $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$, producing the $n\times d$ dimensional projected data matrix to be plotted in each frame of the tour display.

@@

\noindent and $\mathbf{x}_i, i=1,...,n$ is a $p$-dimensional observation in $\mathbf{X}$ and $\mathbf{a}_k, k=1,2 (=d)$ denoting the columns of the projection matrix, $\mathbf{A}=(\mathbf{a}_1, \mathbf{a}_2)$. $\mathbf{A}$ is assummed to be through 0, because the data has been centered on its mean. $\mathbf{x}_i'$ can be considered to be a normal from the projection plane to $\mathbf{x}_i$, and then the norm of this vector gives the orthogonal distance between point and plane. This can be generalized for $d>2$.

The distance can then be used to display a slice tour by highlighting points for which the orthogonal distance is smaller than a selected cutoff value $h$. In 3D, where for each plane there is only a single orthogonal direction, this defines a flat slice of height $2h$. We could use this single direction to slice systematically from one side of the space to the other. It is also possible to think of front and back, some points are closer to the viewer, and some are far, and depth cues could be used. When going beyond 3D, the dimension of the orthogonal space is $>1$, and all sense of direction is lost. In general, there are $(p-d)$ dimensions orthogonal to the projection plane, so using distance from the points to the plane is the simplest approach to generating slices. Using the euclidean distance results in rotation invariant slicing in the orthogonal space, where a "slice" is spherical in the orthogonal subspace, and has radius $h$.

@@

Choosing the slice thickness is a compromise between what feature size can be resolved and the sparseness of the data. As $p$ increases, the relative number of points that are inside a slice of fixed thickness $h$ will decrease. The exact relation depends on the distribution of the data points. To get a best estimate for $h$, given $p$, assume that the points are uniformly distributed in a hypersphere. This is a rotation invariant uniform distribution in $p$ space, and because with sllicing we are mostly interested in hollowness this is more relevant than assuming a multivariate normal distribution.

@@

The equations above assume that the projection plane, and thus the the slice, passes through 0, which will be the mean of centered data. This (center point) is generally a good option for the slice tour, because as the tour progresses the projection plane changes and can catch non-central concavities, too. However it is straightforward generalize the equations to use any center point, $\mathbf{c}$.

In general $\mathbf{c}$ can be any point in the $p$-dimensional parameter space, but we are only interested in the orthogonal component, the part of the vector extedning out of the projection plane, $\mathbf{c}' = \mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{c}\cdot \mathbf{a}_2 )\mathbf{a}_2$. The generalized measure of orthogonal distance is then
\begin{equation}
v_i^2 = ||\mathbf{x}_i' - \mathbf{c}'||^2 = \mathbf{x}_i'^2 + \mathbf{c}'^2 - 2 \mathbf{x}_i'\cdot\mathbf{c}'
\end{equation}
where the cross term can be expressed as
\begin{equation}
\mathbf{x}_i'\cdot\mathbf{c}' = \mathbf{x}_i\cdot\mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) (\mathbf{x}_i\cdot \mathbf{a}_1) - (\mathbf{c}\cdot \mathbf{a}_2) (\mathbf{x}_i\cdot \mathbf{a}_2).
\end{equation}

Using the generalized distance measure with a cutoff volume, $\epsilon$, then corresponds to moving a slice of fixed thickness, corresponding to a neighborhood of the projection plane through the anchor point $\mathbf{c}$. In 3D, this simply corresponds to moving up or down along the orthogonal direction. Note that selecting an off-center anchor point may result in fewer points inside a slice for most data.

@@

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.4\textwidth]{diagrams/centered-slice.pdf}
\hspace{5mm}
\includegraphics[width=0.4\textwidth]{diagrams/orthogonal-distance.pdf}}
\caption{Illustrations of slicing, through a 3D sphere (left), and deomnstrating the calculation of the orthogonal distance (right). 3D slicing, can be done by sliding the projection plane along the orthogonal direction: centered at the origin (green) and one off-center at $\mathbf{c}$ (blue). This intuition does not transfer to higher dimensions, and it is best to use orthogonal distance between point and projection plane for computing the slice. Because a projection plane has no specific location, for slicing we can prescribe this, as through the data center, or any other point, $\mathbf{c}$.}
\label{fig:diagrams}
\end{figure*}

@@

The slice tour has been implemented in R [@rref] as a new display method `display_slice` in the `tourr` package [@tourr]. In addition to usual parameters the user can choose the volume parameter $\epsilon$ by setting `eps` and, if required, the center point $\mathbf{c}$ is set by the `anchor` argument. By default `eps=0.001` and `anchor=NULL`, resulting in slicing through the mean In addition the user can select the marker symbols for point inside and outside the slice. By default points in the slice are highlighted as `pch_slice = 20` and `pch_other = 46`, i.e. plotting a bullet for points inside the slice, and a dot for points outside. Below we show example code for displaying slices through a hollow 3D sphere.

@@

It is especially instructive to look at the sliced spheres with an off-center anchor point. In this case the different views obtained in the slice tour reveal more of the concave structure of the distribution. The first row of Figure \ref{fig:3anchored} shows example views from a slice tour on a 3D sphere with $R=1$ and anchor point $(0.7, 0.7, 0.7)$. Notice that the anchor point was chosen to fall outside the sphere. Depending on the viewing angle selected by the tour the slice contains points on the circle with radius $R$ (when the anchor point has a negligible orthogonal component to the viewing plane); a circle with radius $<R$ as the viewing angle is tilted away from the axis connecting the anchor point to the origin; a full circle with small radius as the angle increases; and finally we see an empty slice when the projection plane is orthogonal to the anchor point axis. A similar picture is found for the 5D sphere, see second row of Figure \ref{fig:3anchored}. For this example we generate 20k points on a hollow 5D sphere to resolve the features. Since $h$ increases with $p$ the resolution is reduced compared to the 3D example. Note also that as dimensionality increases, the larger orthogonal space means that the center point will have a large orthogonal component in most views.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored/sphere-3-anchored-033.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored/sphere-3-anchored-036.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored/sphere-3-anchored-041.png}}
\centerline{\includegraphics[width=0.2\textwidth]{gifs/png/sphere-5-anchored/sphere-5-anchored-014.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-5-anchored/sphere-5-anchored-010.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-5-anchored/sphere-5-anchored-001.png}}
\caption{Different slices through a 3D (first row) and 5D (second row) hollow sphere with $R=1$, with shifted center point, showing full circle with small radius (left), circle with radius $<R$ (middle) and circle with radius $R$ (right).}
\label{fig:3anchored}
\end{figure*}

@@

### Needle in a haystack
We use the pollen data as an example for a hidden feature generally occluded in projections. This is a classic 5D data set, originally simulated by by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen].  The standardised data is observed in a slice tour with a thin slice ($\epsilon = 0.0005$). Selected views are shown in the first two plots in Figure \ref{fig:pollen}. They indicate the presence of an interesting feature hidden in the center, which can be identified as the word "EUREKA" by zooming.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.2\textwidth]{gifs/png/pollen/pollen-007.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/pollen/pollen-092.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/pollen-zoom/pollen-zoom-075.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/pollen-zoom/pollen-zoom-018.png}
}
\caption{Different slices through the classic pollen dataset. The first two plots have slice volume $\epsilon = 0.0005$, the last two plots zoom in on the center and have increased volume $\epsilon = 0.005$. The hidden word can be seen in the zoomed slice.}
\label{fig:pollen}
\end{figure*}

### Non-linear boundaries
@sam.11271 described some principles and approaches for visualising models in the data space. Much of this is based on examining projections provided by a tour of the model in the high-dimensional space. Classification boundaries are explored for the wine data set [@asuncion:2007]. It is difficult to digest the boundaries fully -- for example, where one group's boundary wraps another for example, if they ar linear or nonlinear, or whether the boundary goes through the space or is only carving out a corner of it. The sliced projections makes this easier.

Selected views of projections and sliced projections from a radial basis SVM on 3 variables, and a polynomial basis SVM on 5 variables are shown in Figure \ref{fig:wine}. A large slice volume $\epsilon = 0.15$ is used. The slicing allows exploring the center of the space. With 3D it reveals the spherical boundary of the group (green) hidden by the projection. In 5D with the polynomial basis, projection might suggest that the boundary between classes is almost linear, but the slicing shows that it to be nonlinear near the center.

>>>>

<!-- Background -->
Data is commonly high-dimensional, and visualisation often relies on some form of dimension reduction. This can be done by taking linear projections, or nonlinear if one considers techniques like multidimensional scaling (MDS) [@mds] or t-Distributed Stochastic Neighbour Embedding (t-SNE) [@tsne]. For the purposes here, the focus is on linear projections, in particular as provided by the grand tour [@As85; @BCAH05]. Interactive and dynamic displays can provide information beyond what can be achieved in a static display. The grand tour shows a smooth sequence of interpolated low dimensional projections, and allows the viewer to extrapolate from the low-dimensional shapes to the multidimensional distribution. It is particularly useful for detecting clusters, outliers and non-linear dependence.

<!-- The disadvantage of projection: opacity, occlusion, even with pseudo-transparency from scatterplots;  motivation -->
A major limitation of projections is their opacity. It is mitigated some, when scatterplots are used to render the projected data, through a pseudo-transparency of sparseness of points. This is improved further if points are also drawn using an alpha level that provides transparent dots on most display devices. Some features of multivariate distributions may be visible, but a lot is easily hidden, especially in the case of concave structures. Considering the example of a simple geometric shape such as a hypershere, it is difficult to distinguish between a full or a hollow sphere, that is, whether points are uniformly distributed within the sphere or on the surface of a sphere. Similarly we can think of small scale structures hidden in the centre of a multivariate distribution, that might be considered to be "needles in a haystack", which can be difficult to detect in projections. Projections also obscure non-linear boundaries as might be constructed from a classification model, or nonlinear model fits in high-dimensions.

@@

For the explanation of the slice tour, the actual mechanics of the tour are not important, and only the notion of a projection plane in high dimensions is needed. The notation used for explanations starts with denoting each projection as $\mathbf{X}$ the $n\times p$ dimensional data matrix ($n$ observations in $p$ dimensions) and $\mathbf{A}$ an orthonormal $p \times d$ projection matrix. The $d$-dimensional projection of the data is thus given by $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$, producing the $n\times d$ dimensional projected data matrix to be plotted in each frame of the tour display.

@@

\noindent and $\mathbf{x}_i, i=1,...,n$ is a $p$-dimensional observation in $\mathbf{X}$ and $\mathbf{a}_k, k=1,2 (=d)$ denoting the columns of the projection matrix, $\mathbf{A}=(\mathbf{a}_1, \mathbf{a}_2)$. $\mathbf{A}$ is assumed to be through 0, because the data has been centred on its mean. $\mathbf{x}_i'$ can be considered to be a normal from the projection plane to $\mathbf{x}_i$, and then the norm of this vector gives the orthogonal distance between point and plane. This can be generalized for $d>2$.

The distance can then be used to display a slice tour by highlighting points for which the orthogonal distance is smaller than a selected cutoff value $h$. In 3D, where for each plane there is only a single orthogonal direction, this defines a flat slice of height $2h$. We could use this single direction to slice systematically from one side of the space to the other. It is also possible to think of front and back, some points are closer to the viewer, and some are far, and depth cues could be used. When going beyond 3D, the dimension of the orthogonal space is $>1$, and all sense of direction is lost. In general, there are $(p-d)$ dimensions orthogonal to the projection plane, so using distance from the points to the plane is the simplest approach to generating slices. Using the Euclidean distance results in rotation invariant slicing in the orthogonal space, where a "slice" is spherical in the orthogonal subspace, and has radius $h$.

@@

Choosing the slice thickness is a compromise between what feature size can be resolved and the sparseness of the data. As $p$ increases, the relative number of points that are inside a slice of fixed thickness $h$ will decrease. The exact relation depends on the distribution of the data points. To get a best estimate for $h$, given $p$, assume that the points are uniformly distributed in a hypersphere. This is a rotation invariant uniform distribution in $p$ space, and because with slicing we are mostly interested in hollowness this is more relevant than assuming a multivariate normal distribution.

@@

The equations above assume that the projection plane, and thus the the slice, passes through 0, which will be the mean of centred data. This (centre point) is generally a good option for the slice tour, because as the tour progresses the projection plane changes and can catch non-central concavities, too. However it is straightforward generalize the equations to use any centre point, $\mathbf{c}$.

In general $\mathbf{c}$ can be any point in the $p$-dimensional parameter space, but we are only interested in the orthogonal component, the part of the vector extending out of the projection plane, $\mathbf{c}' = \mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{c}\cdot \mathbf{a}_2 )\mathbf{a}_2$. The generalized measure of orthogonal distance is then
\begin{equation}
v_i^2 = ||\mathbf{x}_i' - \mathbf{c}'||^2 = \mathbf{x}_i'^2 + \mathbf{c}'^2 - 2 \mathbf{x}_i'\cdot\mathbf{c}'
\end{equation}
where the cross term can be expressed as
\begin{equation}
\mathbf{x}_i'\cdot\mathbf{c}' = \mathbf{x}_i\cdot\mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) (\mathbf{x}_i\cdot \mathbf{a}_1) - (\mathbf{c}\cdot \mathbf{a}_2) (\mathbf{x}_i\cdot \mathbf{a}_2).
\end{equation}

Using the generalized distance measure with a cutoff volume, $\epsilon$, then corresponds to moving a slice of fixed thickness, corresponding to a neighbourhood of the projection plane through the anchor point $\mathbf{c}$. In 3D, this simply corresponds to moving up or down along the orthogonal direction. Note that selecting an off-centre anchor point may result in fewer points inside a slice for most data.

@@

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.4\textwidth]{diagrams/centered-slice.pdf}
\hspace{5mm}
\includegraphics[width=0.4\textwidth]{diagrams/orthogonal-distance.pdf}}
\caption{Illustrations of slicing, through a 3D sphere (left), and demonstrating the calculation of the orthogonal distance (right). 3D slicing, can be done by sliding the projection plane along the orthogonal direction: centred at the origin (green) and one off-centre at $\mathbf{c}$ (blue). This intuition does not transfer to higher dimensions, and it is best to use orthogonal distance between point and projection plane for computing the slice. Because a projection plane has no specific location, for slicing we can prescribe this, as through the data centre, or any other point, $\mathbf{c}$.}
\label{fig:diagrams}
\end{figure*}

@@

The slice tour has been implemented in R [@rref] as a new display method `display_slice` in the `tourr` package [@tourr]. In addition to usual parameters the user can choose the volume parameter $\epsilon$ by setting `eps` and, if required, the centre point $\mathbf{c}$ is set by the `anchor` argument. By default `eps=0.001` and `anchor=NULL`, resulting in slicing through the mean In addition the user can select the marker symbols for point inside and outside the slice. By default points in the slice are highlighted as `pch_slice = 20` and `pch_other = 46`, i.e. plotting a bullet for points inside the slice, and a dot for points outside. Below we show example code for displaying slices through a hollow 3D sphere.

@@

It is especially instructive to look at the sliced spheres with an off-centre anchor point. In this case the different views obtained in the slice tour reveal more of the concave structure of the distribution. The first row of Figure \ref{fig:3anchored} shows example views from a slice tour on a 3D sphere with $R=1$ and anchor point $(0.7, 0.7, 0.7)$. Notice that the anchor point was chosen to fall outside the sphere. Depending on the viewing angle selected by the tour the slice contains points on the circle with radius $R$ (when the anchor point has a negligible orthogonal component to the viewing plane); a circle with radius $<R$ as the viewing angle is tilted away from the axis connecting the anchor point to the origin; a full circle with small radius as the angle increases; and finally we see an empty slice when the projection plane is orthogonal to the anchor point axis. A similar picture is found for the 5D sphere, see second row of Figure \ref{fig:3anchored}. For this example we generate 20k points on a hollow 5D sphere to resolve the features. Since $h$ increases with $p$ the resolution is reduced compared to the 3D example. Note also that as dimensionality increases, the larger orthogonal space means that the centre point will have a large orthogonal component in most views.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored/sphere-3-anchored-033.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored/sphere-3-anchored-036.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored/sphere-3-anchored-041.png}}
\centerline{\includegraphics[width=0.2\textwidth]{gifs/png/sphere-5-anchored/sphere-5-anchored-014.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-5-anchored/sphere-5-anchored-010.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-5-anchored/sphere-5-anchored-001.png}}
\caption{Different slices through a 3D (first row) and 5D (second row) hollow sphere with $R=1$, with shifted centre point, showing full circle with small radius (left), circle with radius $<R$ (middle) and circle with radius $R$ (right).}
\label{fig:3anchored}
\end{figure*}

@@

### Needle in a haystack
We use the pollen data as an example for a hidden feature generally occluded in projections. This is a classic 5D data set, originally simulated by by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen].  The standardised data is observed in a slice tour with a thin slice ($\epsilon = 0.0005$). Selected views are shown in the first two plots in Figure \ref{fig:pollen}. They indicate the presence of an interesting feature hidden in the centre, which can be identified as the word "EUREKA" by zooming.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.2\textwidth]{gifs/png/pollen/pollen-007.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/pollen/pollen-092.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/pollen-zoom/pollen-zoom-075.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/pollen-zoom/pollen-zoom-018.png}
}
\caption{Different slices through the classic pollen dataset. The first two plots have slice volume $\epsilon = 0.0005$, the last two plots zoom in on the centre and have increased volume $\epsilon = 0.005$. The hidden word can be seen in the zoomed slice.}
\label{fig:pollen}
\end{figure*}

### Non-linear boundaries
@sam.11271 described some principles and approaches for visualising models in the data space. Much of this is based on examining projections provided by a tour of the model in the high-dimensional space. Classification boundaries are explored for the wine data set [@asuncion:2007]. It is difficult to digest the boundaries fully -- for example, where one group's boundary wraps another for example, if they are linear or nonlinear, or whether the boundary goes through the space or is only carving out a corner of it. The sliced projections makes this easier.

Selected views of projections and sliced projections from a radial basis SVM on 3 variables, and a polynomial basis SVM on 5 variables are shown in Figure \ref{fig:wine}. A large slice volume $\epsilon = 0.15$ is used. The slicing allows exploring the centre of the space. With 3D it reveals the spherical boundary of the group (green) hidden by the projection. In 5D with the polynomial basis, projection might suggest that the boundary between classes is almost linear, but the slicing shows that it to be nonlinear near the centre.
